Using spatially biased variables in ecosystem condition accounting with a GIS based workflow

### 2024-01-16 manuscript/HIAs.qmd
use defult citation style (numbers)

---
title: Using spatially biased variables in ecosystem condition accounting with a GIS based workflow
subtitle: A Short Subtitle
author:
  - name: Anders Lorentzen Kolstad
    email: anders.kolstad@nina.no
    affiliations: 
        - id: some-tech
          name: Norwegian Institute for Nature Research
          department: Department of Terrestrial Ecology
          address: Street Address
          city: City
          state: State
          postal-code: Postal Code
    attributes:
        corresponding: true
    note: This is the first author footnote.
  - name: Bob Security
    email: bob@example.com
    affiliations:
        - id: another-u
          name: Another University
          department: Department Name
          address: Street Address
          city: City
          state: State
          postal-code: Postal Code
    note: |
      Another author footnote, this is a very long footnote and it should be a really long footnote. But this footnote is not yet sufficiently long enough to make two lines of footnote text.
  - name: Cat Memes
    email: cat@example.com
    affiliations:
        - ref: another-u
    note: Yet another author footnote.
  - name: Derek Zoolander
    email: derek@example.com
    affilations:
        - ref: some-tech
abstract: |
  This is the abstract. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Vestibulum augue turpis, dictum non malesuada a, volutpat eget velit. Nam placerat turpis purus, eu tristique ex tincidunt et. Mauris sed augue eget turpis ultrices tincidunt. Sed et mi in leo porta egestas. Aliquam non laoreet velit. Nunc quis ex vitae eros aliquet auctor nec ac libero. Duis laoreet sapien eu mi luctus, in bibendum leo molestie. Sed hendrerit diam diam, ac dapibus nisl volutpat vitae. Aliquam bibendum varius libero, eu efficitur justo rutrum at. Sed at tempus elit.
keywords: 
  - SEEA EA
  - alien species
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Ecological Indicators
      formatting: preprint
      model: 3p
      cite-style: super
---
# Introduction

@@

@@

A general requirement for indicators in the SEEA EA framework is that they should give an unbiased representation of the condition inside the ecosystem assets (Czucz et al 2021; Table 1 - validity; see also SEEA EA §2.87) Ecosystem assets are defined as "ecological entities [meaning areas] about which information is sought and about which statistics are ultimately compiled (REF). This requirement for indicator validity means that spatially biased data are ill suited, especially if sampling intensity varies along gradient of anthropogenic pressures and hence ecosystem condition. SEEA EA is spatially explicit, which in practice means that indicators that are in some way sampled (i.e. not complete wall-to-wall data like remotely sensed imagery), the values are projected on to areas on the map, so that all areas inside the ecosystem accounting area get assigned value for that indicator. There are at least 3 general ways to achieve this complete areal coverage of indicator values:

@@

To describe these two concepts we use five condition variables from the same data set (see @tbl-variables.

| Variable code        | Variable name                                    | Measurement unit                                                                        | Description                                                                                                    | Reference  |
|---------------|---------------|---------------|---------------|---------------|
| 7FK                  | Prevalence of alien species                      | Unitless, ordinal, non-linear scale from 1 (no alien species) to 7 (only alien species) | The fraction of the species composition made up from alien species                                             | Halvorsen  |
| 7SE                  | Human caused abration or abration-caused erosion | Unitless, ordinal, non-linear scale from 1 to 4.                                        | Measures the frequency of imagined 4 m^2^ quadrats layed over the area that has some sign of abration          | ref        |
| PRSL                 | *as above*                                       | Unitless, ordinal, non-linear scale from 0 to 7.                                        | Same as 7SE, but recorded at a higher resolution                                                               | Mdir       |
| 7TK                  | Tracks from large vehicles                       | Unitless, ordinal, non-linear scale from 1 to 4.                                        | Measures the frequency of imagined 100 m^2^ quadrats layed over the area that has some signs of vehicle tracks | Halvorsen  |
| PRTK                 | *as above*                                       | Unitless, ordinal, non-linear scale from 0 to 7.                                        | Same as 7TK, but recorded at a higher resolution                                                               | Mdir       |
| Infrastructure Index | Infrastructure Index                             | Unitelss linear scale from 0 to 13.2                                                    | Unitelss index ranging from from 0 to 13.2                                                                     | Erikstad   |
| HIA                  | Human Impact area                                | Ordinal, non-linear scale from 1 to 4                                                   | A categorical representation of the infrastructure index                                                       | this paper |

Table: Variables used in this study {#tbl-variables}

Reference to @fig-carsplot

```{r fig-carsplot, fig.cap="Test caption"}
plot(cars)
```

@@

# Data availability {-}

>>>>

---
title: Using spatially biased variables in ecosystem condition accounting with a GIS based workflow
subtitle: A Short Subtitle
author:
  - name: Anders Lorentzen Kolstad
    email: anders.kolstad@nina.no
    affiliations: 
        - id: some-tech
          name: Norwegian Institute for Nature Research
          department: Department of Terrestrial Ecology
          address: Street Address
          city: City
          state: State
          postal-code: Postal Code
    attributes:
        corresponding: true
    note: This is the first author footnote.
  - name: Bob Security
    email: bob@example.com
    affiliations:
        - id: another-u
          name: Another University
          department: Department Name
          address: Street Address
          city: City
          state: State
          postal-code: Postal Code
    note: |
      Another author footnote, this is a very long footnote and it should be a really long footnote. But this footnote is not yet sufficiently long enough to make two lines of footnote text.
  - name: Cat Memes
    email: cat@example.com
    affiliations:
        - ref: another-u
    note: Yet another author footnote.
  - name: Derek Zoolander
    email: derek@example.com
    affilations:
        - ref: some-tech
abstract: |
  This is the abstract. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Vestibulum augue turpis, dictum non malesuada a, volutpat eget velit. Nam placerat turpis purus, eu tristique ex tincidunt et. Mauris sed augue eget turpis ultrices tincidunt. Sed et mi in leo porta egestas. Aliquam non laoreet velit. Nunc quis ex vitae eros aliquet auctor nec ac libero. Duis laoreet sapien eu mi luctus, in bibendum leo molestie. Sed hendrerit diam diam, ac dapibus nisl volutpat vitae. Aliquam bibendum varius libero, eu efficitur justo rutrum at. Sed at tempus elit.
keywords: 
  - SEEA EA
  - alien species
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Ecological Indicators
      formatting: preprint
      model: 3p
    include-in-header:
      text: '\usepackage{lineno}\linenumbers'
editor: 
  markdown: 
    wrap: sentence
---

# Introduction

@@

cite-style: super

@@

A general requirement for indicators in the SEEA EA framework is that they should give an unbiased representation of the condition inside the ecosystem assets (Czucz et al 2021; Table 1 - validity; see also SEEA EA §2.87) Ecosystem assets are defined as "ecological entities \[meaning areas\] about which information is sought and about which statistics are ultimately compiled (REF). This requirement for indicator validity means that spatially biased data are ill suited, especially if sampling intensity varies along gradient of anthropogenic pressures and hence ecosystem condition. SEEA EA is spatially explicit, which in practice means that indicators that are in some way sampled (i.e. not complete wall-to-wall data like remotely sensed imagery), the values are projected on to areas on the map, so that all areas inside the ecosystem accounting area get assigned value for that indicator. There are at least 3 general ways to achieve this complete areal coverage of indicator values:

@@

To describe these two concepts we use five condition variables from the same data set ( @tbl-variables).

| Variable code        | Variable name                                    | Measurement unit                                                                        | Description                                                                                                    | Reference                          |
|---------------|---------------|---------------|---------------|---------------|
| 7FK                  | Prevalence of alien species                      | Unitless, ordinal, non-linear scale from 1 (no alien species) to 7 (only alien species) | The fraction of the species composition made up from alien species                                             | @rune_halvorsen_dokumentasjon_2019 |
| 7SE                  | Human caused abration or abration-caused erosion | Unitless, ordinal, non-linear scale from 1 to 4.                                        | Measures the frequency of imagined 4 m^2^ quadrats layed over the area that has some sign of abration          | ref                                |
| PRSL                 | *as above*                                       | Unitless, ordinal, non-linear scale from 0 to 7.                                        | Same as 7SE, but recorded at a higher resolution                                                               | Mdir                               |
| 7TK                  | Tracks from large vehicles                       | Unitless, ordinal, non-linear scale from 1 to 4.                                        | Measures the frequency of imagined 100 m^2^ quadrats layed over the area that has some signs of vehicle tracks | Halvorsen                          |
| PRTK                 | *as above*                                       | Unitless, ordinal, non-linear scale from 0 to 7.                                        | Same as 7TK, but recorded at a higher resolution                                                               | Mdir                               |
| Infrastructure Index | Infrastructure Index                             | Unitelss linear scale from 0 to 13.2                                                    | Unitelss index ranging from from 0 to 13.2                                                                     | Erikstad                           |
| HIA                  | Human Impact area                                | Ordinal, non-linear scale from 1 to 4                                                   | A categorical representation of the infrastructure index                                                       | this paper                         |

: Variables used in this study {#tbl-variables}

@@

# Data availability {.unnumbered}
